Network Origins of Social Distinction

### 2025-07-18 manuscript.tex
Updates from Overleaf

\documentclass[12pt]{article}
\usepackage[left=1.15in, right=1.15in, top=1in, bottom=1.5in]{geometry}
\usepackage{natbib}
\usepackage{verbatim}
\usepackage{setspace}
\usepackage{booktabs}
\usepackage{caption}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage[hidelinks]{hyperref}
\bibpunct[: ]{(}{)}{;}{a}{}{,} 
\usepackage{authblk}
\renewcommand\Authfont{\normalsize}
\renewcommand\Affilfont{\footnotesize}

@@

In this last respect, the more encompassing notion of \textit{social} distinction, implying an actor's embeddedness in a relational field \citep{erikson2013formalist-400, fuhse2021social-9cb}, is often measured as the shadow counterpart of cultural distinction or collapsed into the concept of status more generally \citep{podolny2010status-1a6}. This approach overlooks questions of how network constraints impact the capacity of actors to be distinct. A generalized, measured ordering of social distinction that is derivative from a network of concrete social relations would give Bourdieu's generalized conception of distinction a broader methodological base from which to conduct empirical analysis.

@@

The second aspect of social distinction is \textit{autonomy}, which is the absence of network constraint. Being structurally unique minimizes the social constraints placed on an individual. The tighter the interconnections of the group, the more its pressures constrain the individual. In politics, this aspect of distinction takes the form of constraint. When an actor is one of many densely interconnected individuals, their ability to act contrary to their networked group is greatly weakened. However, when one is only a weakly dependent and connected member of a group, they have a greater ability to operate independently. Their ability to act is a source of distinction as they are able to take independent action. The question of structural uniqueness and autonomy is the same underlying consideration as captured by measures of path centrality. The more critical one is to bridge connections between individuals within a network, the greater one's importance, and the lower their constraint \citep{burt92}. \citet[1263-4]{padgett_ansell94} were the first to bring attention to how the cohesion of those around a broker can structure their own decisions. Cosimo de Medici, acting as a broker between two antagonistic groups, was constrained in the actions he could take because strong actions favoring one side would upset the other side. However, such a constraint was not an issue because Medici’s interests were in perpetuating his central role as broker, and, as the only source of interest, he was able to maintain the position. With regard to conceptualizing distinction, the theory of robust action puts forward that those at the center of political networks, as long as their interests derive solely from maintaining that position, engage in moderate strategies that do not overly antagonize either side. The decisions derive from ensuring the continuation of their position at the center of the network.

@@

\section{A Generalized Measure of Network Distinction}
Below, we present a measure of distinction centrality, determine the appropriate application of a scalar factor, and indicate the availability of an R program to facilitate easy implementation of the measure in further wok.

@@

Let $\beta_i$ be the distinction centrality of a given individual \textit{i}. The first term captures the status of a given actor. Let \textit{x} be the eigenvector centrality \citep{bonacich72} of the designated individual. Let S be a scalar that adjusts the social value placed on status relative to constraint, as discussed below. The second terms captures the element of autonomy. Removing individual \textit{i} from the network, compute the eigenvectors of the remaining individuals and sum those that are connected to individual \textit{i}, taking the average eigenvector of neighbors \textit{j} as the constraint confronting node \textit{i} in the network. This is substantively the same mechanism as conducted in a jackknife resampling procedure, except that rather than forming a sampling distribution, we are measuring local constraint. Finally, subtract the status of the average neighbor connected to \textit{i} from \textit{i}’s scalar adjusted eigenvector centrality in the whole network.

If a node is an isolate, its eigenvector centrality should be set at 0. In the case of a network with no connections (such as a star graph), it can be accidentally calculated as 1.00 in programs. However, given that it is still a network with no connections, setting isolates at 0.00 is theoretically appropriate, considering that status is a reflection of connections. It would be inappropriate for an isolate to have the same status as the highest status node in a large network. Additionally, given the second step of subtraction, it is only theoretically meaningful if set as 0 in the case of distinction centrality.

@@

While an arbitrary scalar is possible, it can yield erroneous results when applied to empirical networks. However, a scalar can be derived from the network itself due to the field positional interplay between status and constraint. Imagining two actors of equal status and personal constraint, one in an averagely more constrained system and the other in a less averagely constrained system, with average status otherwise equal. In such a case, the status of an actor in the more constrained system has a greater impact on their social distinction than in the less constrained system. With high constraint, status takes on greater importance in determining social distinction because it compels deference. In other words, it is the difference between a hierarchical system with and without constraints. Status with a weak constraint means a weaker compulsion of deference. The structure of constraint redounds into the impact of status on social distinction beyond its contribution.

@@

\subsection{Implementation}
To conduct the analyses below, we use a custom function written for the \textit{R} statistical computing environment \citep{R} software.\footnote{The function can be found here \href{https://github.com/olizardo/Jilbert-Distinction-Centrality/blob/main/distinction.R}{https://github.com/olizardo/Jilbert-Distinction-Centrality/blob/main/distinction.R}}

@@

The paper was an early application of an algorithm to predict the lines along which the group would divide. The original Ford-Fulkerson Labeling algorithm was a network flow model which sought to use the length between nodes to detect community membership (1977:454). However, in the algorithm, Node 9 was mis-assigned to the Officer’s network using the original algorithm, whereas empirically Node 9 joined the network of the karate teacher, Mr. Hi. The original discrepancy was explained in reference to Node 9 being near to achieving a black belt, and so the institutional opportunity costs overrode the social network costs.

@@

The exercise, however, focuses attention on the boundaries of the field. The original explanation for the discrepancy points to Node 9's large karate capital. If the field is bounded as the ``karate club,'' Node 9's decision is unexpected, but if the field is ``karate practitioners'' then there might be much of the field that is going unobserved in Zachary's original study. We do know that some of the students, including Node 9, practiced at Mr. Hi's dojo as well as the club. The fact that Node 9 had club connections only to the most central members of both the officer's and Mr. Hi's networks would indicate that Node 9 only affiliated with likely similarly advanced members of the club. If Node 9 was similarly affiliating with advanced students who only attended the dojo, then this would likely influence the decision regarding distinction at the field level. Failure to completely capture the positions in the appropriately defined field can create measures that fail to capture empirical outcomes.

@@

\section{Distinction with Structural Holes vs. Structural Folds}

@@

@@

>>>>

\documentclass[12pt]{article}
\usepackage[left=1.15in, right=1.15in, top=1in, bottom=1.5in]{geometry}
\usepackage{natbib}
\usepackage{verbatim}
\usepackage{setspace}
\usepackage{booktabs}
\usepackage{caption}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{listings}
\usepackage{courier}
\lstset{basicstyle=\footnotesize\ttfamily,breaklines=true}
\lstset{framextopmargin=50pt,frame=bottomline}
\usepackage[hidelinks]{hyperref}
\bibpunct[: ]{(}{)}{;}{a}{}{,} 
\usepackage{authblk}
\renewcommand\Authfont{\normalsize}
\renewcommand\Affilfont{\footnotesize}

@@

In this last respect, the more encompassing notion of \textit{social} distinction, implying an actor's embeddedness in a relational field \citep{erikson2013formalist-400, fuhse2021social-9cb}, is often measured as the shadow counterpart of cultural distinction or collapsed into the concept of status more generally \citep{podolny2010status-1a6}. This approach overlooks questions of how network constraints produced by other agents impact the capacity of actors to be distinct. A generalized, measured ordering of social distinction that is derivative from a network of concrete social relations would give Bourdieu's generalized conception of distinction a broader methodological base from which to conduct empirical analysis.

@@

The second aspect of social distinction is \textit{autonomy}, which is the absence of network constraint \citep{burt1980autonomy}. Being structurally unique minimizes the social constraints placed on an individual. The tighter the interconnections of the group, the more its pressures constrain the individual. In politics, this aspect of distinction takes the form of constraint. When an actor is one of many densely interconnected individuals, their ability to act contrary to their networked group is greatly weakened. However, when one is only a weakly dependent and connected member of a group, they have a greater ability to operate independently. Their ability to act is a source of distinction as they are able to take independent action. The question of structural uniqueness and autonomy is the same underlying consideration as captured by measures of path centrality. The more critical one is to bridge connections between individuals within a network, the greater one's importance, and the lower their constraint \citep{burt1995structural-5e8}.

\citet[1263-4]{padgett1993robust} were the first to bring attention to how the cohesion of those around a broker can structure their own decisions. Cosimo de Medici, acting as a broker between two antagonistic groups, was constrained in the actions he could take because strong actions favoring one side would upset the other side. However, such a constraint was not an issue because Medici’s interests were in perpetuating his central role as broker, and, as the only source of interest, he was able to maintain the position. With regard to conceptualizing distinction, the theory of robust action puts forward that those at the center of political networks, as long as their interests derive solely from maintaining that position, engage in moderate strategies that do not overly antagonize either side. The decisions derive from ensuring the continuation of their position at the center of the network.

@@

\section{A Generalized Measure of Network Distinction}
Below, we present a measure of distinction centrality, determine the appropriate application of a scalar factor, and indicate the availability of an R program to facilitate easy implementation of the measure in further work.

@@

Let $\beta_i$ be the distinction centrality of a given individual \textit{i}. The first term captures the status of a given actor. Let \textit{x} be the eigenvector centrality of the designated individual. Let S be a scalar that adjusts the social value placed on status relative to constraint, as discussed below. The second term captures the element of autonomy. Removing individual \textit{i} from the network, compute the eigenvectors of the remaining individuals and sum those that are connected to individual \textit{i}, taking the average eigenvector of neighbors \textit{j} as the constraint confronting node \textit{i} in the network. This is substantively the same mechanism as conducted in a jackknife resampling procedure, except that rather than forming a sampling distribution, we are measuring local constraints. Finally, subtract the status of the average neighbor connected to \textit{i} from \textit{i}’s scalar-adjusted eigenvector centrality in the whole network.

When a node is an isolate, we set its eigenvector centrality to $x_i =0$. In the case of a network with no connections (such as an empty graph), some routines calculate it as $x_i =1.0$. However, given that it is still a network with no connections, setting isolates at $x_i =0$ is theoretically appropriate, considering that status is a reflection of connections. It would be inappropriate for an isolate to have the same status as the highest-status node in a large network. Additionally, given the second step of subtraction, it is only theoretically meaningful if set as $x_i =0$ in the case of distinction centrality.

@@

While an arbitrary scalar is possible, it can yield erroneous results when applied to empirical networks. However, a scalar can be derived from the network itself due to the field positional interplay between status and constraint. Imagining two actors of equal status and personal constraint, one in an averagely more constrained system and the other in a less averagely constrained system, with average status otherwise equal. In such a case, the status of an actor in the more constrained system has a greater impact on their social distinction than in the less constrained system. With high constraint, status takes on greater importance in determining social distinction because it compels deference. In other words, it is the difference between a hierarchical system with and without constraints. Status with a weak constraint means a weaker compulsion of deference. The structure of constraint has a reverberative impact on social distinction beyond its contribution.

@@

\subsection{Distinction with Structural Holes vs. Structural Folds}

%Clique addition experiment

%Add cliques to star graph and see what happens

%See what happens with the caveman

%Placehold graphs
%\begin{figure}
%   \captionsetup[subfigure]{font=footnotesize,labelfont=footnotesize}
%    \centering
%     \begin{subfigure}[b]{0.45\textwidth}
%        \includegraphics[width=1.0\textwidth]{Plots/sf.png}
%            \caption{Star Graph}
%            \label{fig:Placehold}
%    \end{subfigure}
%     \begin{subfigure}[b]{0.45\textwidth}
%        \includegraphics[width=1.0\textwidth]{Plots/sf.png}
%            \caption{Structural Fold}
%            \label{fig:Placehold}
%    \end{subfigure}
%    \caption{Structural Holes with more added subgraphs.}
%    \label{fig:toys}
%\end{figure}

\subsection{Long Ties and Hierarchy}
This section highlights how the hierarchy is really formed out of the redundant connections of the subgraphs. While a star graph clearly creates a hierarchy, when additional weight is added to the broker's alters, the hierarchy can be flipped, especially when they form denser subgraphs. This is in keeping with the observed structural holes above.

Graph (b) is the same as graph (a),but with Nodes 12, 13, and 14 combined into a clique with Node 9. Graph (c) is the same as graph (b) with the addition of Nodes 21, 22, and 23 as personal connections of Nodes 12, 13, and 14. Finally, graph (d) is the same as graph (c) with the addition of three long ties across subgraphs connecting only Nodes 17 with 23 and 11 with 18.

As Figure 5 graph (d) demonstrates, when those subgroups become connected redundantly not through the central figure, but through the sparse interconnection of the other members, then Node 9's position of distinction becomes highly elevated.

Thus, there is a U shaped pattern where at low levels of interconnection between sparse subgroups, mediated by a central node, the distinction level of the broker is high. However, as those subgraphs become more densely connected within themselves, the position of the central node is weakened as status accrues more heavily in those local subgraphs. However, when sparse long connections then form to connect the subgraphs, Node 9 achieves a high level of distinction again because the overall constraint in the network increases with the added ties between the subgraphs. The resulting decrease in the scalar works to stratify the distinction distance between Node 9 and other high status competitors, while simultaneously decreasing the constraint forced upon Node 9

\begin{figure}
    \captionsetup[subfigure]{font=footnotesize,labelfont=footnotesize}
    \centering
     \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=1.0\textwidth]{Plots/political1.png}
            \caption{High Subgraph Density, Low Graph Density}
            \label{fig:Placehold}
    \end{subfigure}
     \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=1.0\textwidth]{Plots/political2.png}
            \caption{Broker Given Independent Foundations}
            \label{fig:Placehold}
    \end{subfigure}
         \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=1.0\textwidth]{Plots/political3.png}
            \caption{Increased Independent Foundations}
            \label{fig:Placehold}
    \end{subfigure}
     \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=1.0\textwidth]{Plots/political4.png}
            \caption{All Subgraphs Connected by Long Ties}
            \label{fig:Placehold}
    \end{subfigure}
    \caption{Increasingly connected graphs with central broker.}
    \label{fig:Evolving Political Network}
\end{figure}

\input{Tabs/political.tex}

\subsection{Implementation}
To conduct the analyses below, we use a custom function written for the \textit{R} statistical computing environment \citep{R} software. The function code is reproduced in the \hyperref[sec:appendix]{the Appendix}.

@@

The paper was an early application of an algorithm to predict the lines along which the group would divide. The original Ford-Fulkerson Labeling algorithm was a network flow model that sought to use the length between nodes to detect community membership (1977:454). However, in the original algorithm, Node 9 was misclassified to the Officer’s network, whereas empirically Node 9 joined the network of the karate teacher, Mr. Hi. The original discrepancy was explained in reference to Node 9 being near to achieving a black belt, and so the institutional opportunity costs overrode the social network costs.

@@

The exercise, however, focuses attention on the boundaries of the field. The original explanation for the discrepancy points to Node 9's large karate capital. If the field is bounded as the ``karate club,'' Node 9's decision is unexpected, but if the field is ``karate practitioners'' then there might be much of the field that is going unobserved in Zachary's original study. We do know that some of the students, including Node 9, practiced at Mr. Hi's dojo as well as the club. The fact that Node 9 had club connections only to the most central members of both the officer's and Mr. Hi's networks would indicate that Node 9 only affiliated with likely similarly advanced members of the club. If Node 9 was similarly affiliating with advanced students who only attended the dojo, then this would likely influence the decision regarding distinction at the field level. Failure to completely capture the positions in the appropriately defined field can result in measures that fail to accurately capture empirical outcomes.

@@

Given the value that eigenvector-style centrality measures have had in underpinning the logic of internet search algorithms \citep{kleinberg99, brin_page98, langville2005survey-572}, the application of distinction centrality in the case of a reference network is briefly considered. As the social logic of distinction centrality involves jointly considering the interaction of status and constraint, so in a reference network, the distinction measure will give greater priority to the uniqueness of a node's flow of affiliations.

A toy reference network is presented in Figure 3. The network was designed to contrast status versus distinction within a graph, featuring a clear cluster of nodes with an elevated level of density and tenuously connected to a node with a high degree of popularity external to the main, denser cluster. Such a graph is supposed to represent a contrast between an orthodox set of mutual endorsements and the emergence of a heterodox set of connections.

The resulting distinction measures are presented in Table 6. The results are ranked in order of highest to lowest scaled distinction centrality. Node 8 measures highest in both scaled distinction and status, but immediately after Node 9 and Node 6 switch places in their ordering. Node 6 has the second highest status, but the third highest scaled distinction, while Node 9 goes from third in status to second in scaled distinction. The measure rewards the independent connections Node 9 has with Nodes 13 and 14 over the heavier constraints imposed on Node 6. \footnote{In the absence of Node 8's and its constraint effect on Node 6, Node 6 would become the most distinct and highest status.} Node 10's relative position in the social hierarchy is consistent at fourth in both the status and scaled distinction measures.

\begin{figure}
    \captionsetup[subfigure]{font=footnotesize,labelfont=footnotesize}
    \centering
     \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=1.0\textwidth]{Plots/search.png}
            \caption{Reference Network Example}
            \label{fig:search}
    \end{subfigure}
    \caption{Example of a non-directed network of references, e.g. citation network, internet links, etc.}
    \label{fig:search}
\end{figure}

\input{Tabs/search}

The heterodoxy of Node 5 is elevated into the fifth position of the scaled distinction hierarchy compared to the sixth position it has in the status hierarchy. Node 7 correspondingly falls into the sixth position of the scaled distinction hierarchy from the fifth position in terms of status. However, scaled distinction centrality also elevates Nodes 1, 2, 3, and 4 which buttress 5's popularity. These nodes have the lowest status, yet leap over the hanger's on to the core members of the main orthodox cluster. They are greatly penalized for their high level of constraint, whereas Node's 1, 2, 3, and 4 are empowered by 5's own lower status position. If the goal is to elevate uniqueness, this makes sense as Nodes 11 through 16 are likely to conform to their higher status, and high distinction, connections. Nodes 7, 8, 9, and 10 already implicitly represent their contributions. Additionally, the application of the scalar is important in the final results. Without the scalar, Node 5 would actually be the third ranked node in terms of distinction, leaping Node 6. Such a result may be desirable in technical applications where uniqueness should be privileged over social hierarchy. On the whole, the scaled distinction centrality measure thus works to elevate the uniqueness of a node in search applications.

@@

%College as inherently distinction based institution makes liberals a thing, rather than if you only draw from caves

@@

\newpage
\section*{Appendix: Distinction
Centrality Function} \label{sec:appendix}
\begin{lstlisting}[language=R]
distinction <- function(x) { 
   library(igraph) #calling igraph
   has.labels <- as.numeric(is.null(V(x)$name) == FALSE) #checking for vertex labels
   V <- vcount(x)
   if (has.labels == 0) {
      V(x)$name <- 1:V
      names <- V(x)$name 
      }
   if (has.labels == 1) {
      names <- V(x)$name 
      V(x)$name <- 1:V
      }
   s <- eigen_centrality(x)$vector
   d <- rep(V, 0) #initializing distinction vector
   u <- rep(V, 0) #initializing constraint vector
   for (i in as.character(V(x)$name)) {
      j <- names(neighbors(x, i))
      x.d <- delete_vertices(x, i)
      c <- as.numeric(is_connected(x.d) == TRUE) #checking for connectedness
      e <- as.numeric(ecount(x.d) == 0) #checking for empty
      if (c == 1 & e == 0) { #connected non-empty graph
         s.a <- eigen_centrality(x.d)$vector[j]
         } #end if
      else if (c == 0 & e == 1) { #empty graph
         s.a <- 0
         } #end first else 
      else if (c == 0 & e == 0) { #disconnected non-empty graph
         C <- components(x.d)$membership
         names(C) <- V(x)$name[-as.numeric(i)]
         s.a <- rep(0, length(C))
         for (k in unique(C)) {
            sub.g <- subgraph(x.d, names(which(C == k)))
            if (vcount(sub.g) > 1) {
               s.a[which(C == k)] <- eigen_centrality(sub.g)$vector
               }
            }
         names(s.a) <- names(C)
         s.a <- s.a[j]
         } #end second else
      u[i] <- sum(s.a)/length(s.a) #i's average neighbor centrality in node deleted subgraph
      d[i] <- s[i] - u[i] #i's distinction centrality
   } #end i for loop
   d[is.na(d)] <- 0
   u[is.na(u)] <- 0
   scalar <- mean(u)/mean(s)
   sd <- (s*scalar) - u
   dat <- data.frame(d = d, sd = sd, s = s, u = u, scalar = scalar)
   dat <- round(dat, 4)
   dat <- data.frame(n = names, dat)
   rownames(dat) <- 1:nrow(dat)
   return(dat)
} #end function
\end{lstlisting}
